Review Paper Draft

### 2020-06-25 main.tex
Meeting notes/comments

The mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors \anna{directly} influence population-specific allele frequencies, or vice versa, only that the two may be related. \pb{I find these previous two sentences confusing/contradictory: We assume that environment and genetics are related, but also that environment has no effect on SNPs.}

@@

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. It should be noted that by attempting to reduce the effects of population structure, these methods reduce the effects of environmental confounding by blocking the casual pathway between population structure and observed SNPs \pb{Shouldn't this be: between environment and SNPs?}. \anna{no...}

@@

@@

%------------------------------%
\subsection{LMM-Lasso}
LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p).$$  Typically, $\bZ$ is not observed directly, and a common convention \citep[e.g.,][]{Rakitsch2012} is to replace $\bZ$ with $\bX$. Then, rather than assuming anything about $\bgamma$ directly, let $\bu=\bX\bgamma$ represent the random effect and assume $\boldsymbol{u} \sim N(\boldsymbol{0}, \sigmagg \bK)$. This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$ and thus $\by  \sim N(\bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso then performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}}\bX \bbeta, \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
Using the spectral decomposition of $\bK$ described in equation \ref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization \pb{ONLY WORKS IF $\bX$ IS FULL RANK} \anna{I understand this on paper, but why does it work numerically? Even if $n > p$, svd(K) or eigen(K) gives a U that is $n \times n$... what are the vectors corresponding to the 0-eigenvalues? Toy example...}
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$, $\byT = \bU^T \by$, and $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \boldsymbol{D}^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of substantially fewer parameters: $p + 2$ for LMM-Lasso-Rakitsch, and $p + 3$ for LMM-Lasso-ggmix.

@@

\pb{We should discuss this section; to me, this section comes across as overly complicated and the point is not really coming across}

@@

\begin{figure}[H]
    \centering
    % \includegraphics[scale = 0.9]{figures/fig2b.png}
    \includegraphics[scale = 0.9]{figures/mse_diff_hetero.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for different subpopulation and $\bgamma$ structures: \anna{structureX is 1d linear admixture, $\eta= \xi = 0.5$}}
    \label{fig:big_small_gamma}
\end{figure}

>>>>

The mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors \anna{directly} influence population-specific allele frequencies, or vice versa, only that the two may be related.  \anna{often end up related for complex reasons... the environment and genetics have become associated over time for complicated reasons that we aren't going to get into... air quality isn't changing people's genetics, genetics aren't causing people to move to areas of better/worse air quality}

@@

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. It should be noted that by attempting to reduce the effects of population structure, these methods reduce the effects of environmental confounding by blocking the casual pathway between population structure and observed SNPs.

@@

\anna{typically Z is an incidence matrix... subpops q ... but not the only possibility}

@@

%------------------------------%
\subsection{LMM-Lasso}
LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p).$$  Typically, $\bZ$ is not observed directly, and a common convention \citep[e.g.,][]{Rakitsch2012} is to replace $\bZ$ with $\bX$. Then, rather than assuming anything about $\bgamma$ directly, let $\bu=\bX\bgamma$ represent the random effect and assume $\boldsymbol{u} \sim N(\boldsymbol{0}, \sigmagg \bK)$. This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$ and thus $\by  \sim N(\bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso then performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}}\bX \bbeta, \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
Using the spectral decomposition of $\bK$ described in equation \ref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization \pb{ONLY WORKS IF $\bX$ IS FULL RANK} \anna{I understand this on paper, but why does it work numerically? Even if $n > p$, svd(K) or eigen(K) gives a U that is $n \times n$... what are the vectors corresponding to the 0-eigenvalues? Toy example... PROVIDED U IS N X N}
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$, $\byT = \bU^T \by$, and $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \boldsymbol{D}^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of substantially fewer parameters: $p + 2$ for LMM-Lasso-Rakitsch, and $p + 3$ for LMM-Lasso-ggmix.

@@

\pb{We should discuss this section; to me, this section comes across as overly complicated and the point is not really coming across}
\anna{homogeneous reflects groups that are genetically similar have more similar environmental exposures, heterogeneous says this is not the case. For example... negative lung exposures that are geographic make more sense as a homogeneous exposure but something more behavior-based like smoking behaviors are more plausibly be heterogenous: e.g. smoking in China vs. France - similar behavior, different genetically}

@@

\anna{replace this with inedpendent subpopulations}
\begin{figure}[H]
    \centering
    % \includegraphics[scale = 0.9]{figures/fig2b.png}
    \includegraphics[scale = 0.9]{figures/mse_diff_hetero.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for different subpopulation and $\bgamma$ structures: \anna{structureX is 1d linear admixture, $\eta= \xi = 0.5$}}
    \label{fig:big_small_gamma}
\end{figure}
